

The aim of this project is to
extend the RISC-V RV32I or RV32E base architectures \footnote{
 There is no reason that the ISE would not work with the
 RV64I and RV128I base instruction sets.
 Though the ISE state uses only 32-bit words,
 and would not take advantage of a larger word size without
 further modification.
} 
to accelerate common cryptographic
operations associated with block ciphers and public key cryptography.
It proposes a new Instruction Set Extension (ISE)
which acts analogously to a floating-point co-processor,
but targeted at cryptographic operations.

The ISE is designed to be applicable to a wide range of algorithms
which work on different bit-widths and with different block sizes.
The key pieces of functionality the ISE provides are:

\begin{itemize}
\item Flexible {\em packed arithmetic},
    where multiple values of varying lengths can
    be operated on inside a register in parallel.
\item Memory Scatter/Gather functionality for bytes.
\item Flexible support for word, halfword, byte and nibble permutation.
\item Multi-precision integer arithmetic.
\item Support for accelerating Bit-sliced implementations of cryptographic
    algorithms.
\end{itemize}

We hope that by providing flexible ways to accelerate cryptographic functions,
we can also make implementation of side-channel countermeasures
more efficient.

This document {\em is} a specification for an ISE.
It {\em is not} an implementation guide
\footnote{In fact, we go out of our way to make sure this
ISE is applicable to as many sizes and styles of implementation as possible.}
or a standalone ISA.
